```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Basketball Shots as Spatial Objects {#spatial}

__Note that all the ```R``` code used in this book is accessible on [GitHub](https://github.com/olivierchabot17/ballbook).__

Let's load the basketball court and spatial polygons we've built in the previous chapters.

```{r}
# Load the plot_court() function from the previous chapters
source("code/court_themes.R")
source("code/fiba_court_points.R")
# Load the different zone polygon objects
source("code/zone_polygons.R")

# Load libraries
library(tidyverse) # ggplot and dplyr
library(sf) # working with spatial objects
```

Next, we can load the augmented basketball shots data set we created in Chapter \@ref(shots-data).

```{r}
# Load shot data
shots <- readRDS(file = "data/shots_augmented.rds")
```

```{r full-metrics, tidy=FALSE, echo = FALSE}
knitr::kable(
  head(shots) %>% select(-shot_made_numeric, -dist_meters, -theta_rad),
  digits = 2,
  caption = 'Display the first few shots of augmented data',
  booktabs = TRUE
)
```

Note that we have access to who took the shot, whether they made it or not, and from where on the court they released it. From there, we used the Pythagorean theorem to calculate the shot distance from the center of the hoop and we used trigonometric ratios to calculate the angle from the center line. 

## The Spatial Advantage 

We can convert our augmented shot data to an `sf` object to take advantage of the spatial nature of the data.

```{r}
# convert shots to an sf object
shots_sf <- st_as_sf(shots, coords = c("loc_x", "loc_y"))

# View sf object
shots_sf %>% select(-shot_made_numeric, -dist_meters, -theta_rad)
```

The simple fact that our shots dataframe is now an `sf` object means that we can use the `st_join()` function which will automatically create zone columns based on the location of each shot relative to the polygons we created in Chapter \@ref(zones).

```{r}
# shot_zone_range
shots_sf <- st_join(
  x = shots_sf,
  y = distance_polys
) %>%
  # shot_zone_area
  st_join(
    y = angle_polys
  ) %>%
  # shot_zone_basic
  st_join(
    y = basic_polys
  ) %>%
  # area_value
  st_join(
    y = point_polys
  ) %>%
  # shot_value
  mutate(
    shot_value = ifelse(area_value == "Two-Point Area", 2, 3)
  ) %>%
  # Reorder and only keep relevant variables
  select(player, shot_made_numeric, shot_made_factor,
         dist_feet, dist_meters, theta_deg, theta_rad, shot_value,
         shot_zone_range, shot_zone_area, shot_zone_basic, area_value,
         geometry)
```

The easiest way to test whether the join worked properly would be to randomly select a few shots and plot their different zone labels.

```{r}
set.seed(123) # Always Display the same random shots
sample_shots_sf <- shots_sf %>% 
  # Randomly select 20 shots
  slice_sample(n = 20)
```

```{r, echo = FALSE, fig.cap = 'Check labels for the same20 random shots', out.width='100%', fig.align='center'}

# To arrange plots in a 2x2 grid
library(gridExtra)

p1 <- plot_court() +
  geom_sf_text(data = sample_shots_sf,
               aes(color = shot_zone_range, label = shot_zone_range),
               alpha = 0.8, show.legend = FALSE) +
  labs(title = "Distance Based Zones")

p2 <- plot_court() +
  geom_sf_text(data = sample_shots_sf,
               aes(color = shot_zone_area, label = shot_zone_area),
               alpha = 0.8, show.legend = FALSE) +
  labs(title = "Angle Based Zones")

p3 <- plot_court() +
  geom_sf_text(data = sample_shots_sf,
               aes(color = shot_zone_basic, label = shot_zone_basic),
               alpha = 0.8, show.legend = FALSE) +
  labs(title = "Basic Zones")

p4 <- plot_court() +
  geom_sf_text(data = sample_shots_sf,
               aes(color = factor(shot_value), label = shot_value),
               alpha = 0.8, show.legend = FALSE) +
  labs(title = "Point Value Zones")

# plot 4 plots
grid.arrange(p1, p2, p3, p4, ncol=2, nrow =2)
```

It seems like our joins worked properly. Let's save this new data set so we can easily load it in future chapters.

```{r, eval = FALSE}
# Save the spatial data
saveRDS(shots_sf, file = "data/shots_sf.rds")
```


In the next chapter, we will create our first __shot chart__. How exciting! More specifically, we will try to determine whether the shot locations are spatially randomly distributed or if they seem to cluster.

__Note that all the ```R``` code used in this book is accessible on [GitHub](https://github.com/olivierchabot17/ballbook).__


